---
title       : Workflow
subtitle    : BEFdata + rbefdata 
author      : Claas-Thido Pfaff
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
---

```{r setup, cache = F, include = F, eval = T}
# install.packages("devtools")
require(devtools)
# install_github(username = "befdata", repo = "rbefdata", subdir = "rbefdata")
require(rbefdata)
# install_github("taxize_", "ropensci")
require(taxize)
# install.packages("gdata")
require(gdata)
# set options
options("width"=155)
```

## Install packages

* devtools (devel + install github)

```{r install_devtools, eval = F}
install.packages("devtools")
```

* Trim (leading/trailing spaces)

```{r install_gdata, eval = F}
install.packages("gdata")
```

* taxize (taxonomy tasks)

```{r install_taxize, eval = F}
install_github("taxize_", "ropensci")
```

* rbefdata (interface to BEFdata)

```{r install_rbefdata, eval = F}
remove.packages("rbefdata")
install_github(username = "befdata", repo = "rbefdata", subdir = "rbefdata")
```

---

## List Options (rbefdata)


```{r eval = T}
require(rbefdata)
bef.options()
```

---

## Set Options (rbefdata)



- The URL ([Workshop Portal](http://stage.china.befdata.biow.uni-leipzig.de))

```{r eval = T}
bef.options(url = "http://stage.china.befdata.biow.uni-leipzig.de")
```

- The credentials ([your profile](http://stage.china.befdata.biow.uni-leipzig.de/profile))

```{r eval = T}
bef.options(user_credentials = "BCeqHdtGA7v2MHd15l")
```

---
  
## Get data
  
* Get dataset from [proposal](http://stage.china.befdata.biow.uni-leipzig.de/paperproposals/1) with id #1 

```{r get_dataset, eval = T, cache = T}
datasets = bef.portal.get.datasets_for_proposal(id = 1)
length(datasets)
titles = sapply(datasets, function(x) attributes(x)$title)
titles
dataset = datasets[[2]]
```

---

## Inspect the data

* Inspect the dataset (metadata)
  
```{r get_dataset_metadata, eval = T}
attributes(dataset)$title
names(attributes(dataset))
```

---

## Inspect the data

* About the spces column and their description

```{r get_column_description, eval = T, tidy=FALSE}
names(dataset)
attributes(dataset)$column$description
```

---

## Species names

* Prepare species list

```{r extract_species_list, eval = T}
require(gdata)
species = trim(dataset$species)
unique_species = unique(as.character(species))
head(unique_species, 7)
```

* Load ([taxize](http://ropensci.org/tutorials/taxize_tutorial.html))

```{r require_taxize, eval = F}
require(taxize)
```

---

## Species names

* Taxize offers access to databases like: 
    - Integrated Taxonomic Information System [itis](http://www.itis.gov/) 
    - National Center for Biotechnology Information [ncbi](http://www.ncbi.nlm.nih.gov/) 
    - International Union for Conservation of Nature [iucn](http://www.iucn.org/)

* But that is not all (e.g create phylogeny tree, see [tutorial](http://ropensci.org/tutorials/taxize_tutorial.html))

* Example for species resolving

```{r example_species_name_resolving, eval = T, cache = T, message = F}
species_list = c("Helianthus annuus", "Pinus contort")
query_response = tnrs(query = species_list)[, -c(6:7)]
query_response
```

--- bg:#EEE

## Your turn

* Resolve all the species in vector unique_species

```{r eval = F}
tnrs(query = ...)
```

* Find the entries that need correction (Hint: score!)

---

## A solution 

* Check the unique names of our species list

```{r resolv_our_unique_names, eval = T, message = F, cache = T}
query_response = tnrs(query = unique_species)[, -c(6:7)]
head(query_response,9)[c("submittedName", "acceptedName", "score")]
```

--- 

## A solution 

* Pick the entries that are lower than 1

```{r find_potential_wrong, eval = T, message = F, cache = T}
index_of_need_attention = which(query_response$score < 1)
subset_that_needs_attention = unique(query_response[index_of_need_attention,])
subset_that_needs_attention
```

---

## A solution 

* Pick the entries that are lower 1 and bigger 0.5

```{r find_potential_wrong_two, eval = T, message = F, cache = T}
index_of_need_attention = which(query_response$score < 1 & query_response$score > 0.5)
subset_that_needs_attention = unique(query_response[index_of_need_attention,])
subset_that_needs_attention
```

--- 

## Match back and combine

* Match species to subset submitted names

```{r match_back_corrected_species_one, eval = T, message = F, cache = T}
match_index = match(as.character(species), subset_that_needs_attention$submittedName)
```

*  Create new vector containing the corrections

```{r match_back_corrected_species_two, eval = T, message = F, cache = T}
corrected_names = subset_that_needs_attention$matchedName[match_index]
```

---

## Match back and combine 

* merge function picks from vectors (if else)

```{r merge_function, eval = T, cache = T}
merge_em <- function(base_vec, merge_vec) {
  output = vector()
  for(i in 1:length(base_vec)) {
    if(is.na(base_vec[i])) {
      output[i] = merge_vec[i]
    } else {
      output[i] = base_vec[i]
    }
  }
  return(output)
}
```

* Define base and merge vector

```{r base_and_mergin}
base = corrected_names
mergein = as.character(species)
```

---

## Match back and combine

* create new dataframe

```{r prepare_merge, eval = T}
merged_names = merge_em(base_vec = base, merge_vec = mergein)
corrected_dataset = data.frame(dataset, species_corrected = merged_names)
head(corrected_dataset, 15)
```

---

## Upload dataset

* push back a corrected version to BEFdata

* Please use own title if you execute this step 

```{r upload_corrected, eval = F}
title = "Abundance of Tree and Shrub species in the Comparative Study Plots (CSP) - BEFdata tutorial (species corrected)"
bef.portal.upload.dataset(dataset = corrected_dataset, dataset_title = title)
```

---

## Get higher taxonomic names

* Use taxize to retrive higher taxon names

* Create a new dataframe 
     - corrected unique species names 
     - order
     - family
     - genus

* Can be used to calculate phylogenetic diversity

---

## Get higher taxonomic names

* Example on retrieving classifications

```{r example_higher_names_one, eval = T, message = F}
species_examples = c("Homo sapiens", "Rattus rattus")
```

* The databases can be e.g. [itis](http://www.itis.gov/) or [ncbi](http://www.ncbi.nlm.nih.gov/)

```{r example_higher_names_two, eval = T, message = F, cache = T}
classification_list = classification(species_examples, db = "itis")
classification_list[[1]]
```

---

## Get higher taxonomic names

* Get only parts of intrest

```{r example_higher_names_three, eval = T, message = F, cache = T}
tax_name(query = "Homo Sapiens", get = "genus", db = "ncbi")
tax_name(query = species_examples, get = "family", db = "ncbi")
```

--- bg:#EEE

## Your turn

* Get the order, family and genus from ncbi for unique corrected species
    
* Be nice and only query server once!
    - do not use: tax_name(...)
    - use classification(...)

* Hint: use R apply functions
    
* When finished upload the dataset to the portal!

---

## A solution

* Get: 
    - order, family and genus (unique corrected species use "ncbi")

```{r solution_higher_names, eval = T, tidy = F, cache = T, message = F}
species_list = unique(as.character(corrected_dataset$species_corrected))
classification_species = classification(species_list, db = "ncbi")
```

* Extract (order):

```{r extract, eval = T, cache = T}
extraction_ncbi_order = sapply(classification_species, function(x) if(length(x) != 3){NA}else{subset(x, x$"Rank" == "order")}$"ScientificName") 

```
---

## A solution

* Extract (family):

```{r extract_two, eval = T, cacht =T}
extraction_ncbi_family = sapply(classification_species, function(x) if(length(x) != 3){NA}else{subset(x, x$"Rank" == "family")}$"ScientificName") 
```

* Extract (genus):

```{r extract_three, eval = T, cacht =T}
extraction_ncbi_genus = sapply(classification_species, function(x) if(length(x) != 3){NA}else{subset(x, x$"Rank" == "genus")}$"ScientificName")
```

---

## A solution

* Create data frame:

```{r show_solution_dataset, eval = T, cache = T, message = F}
extraction_ncbi_order_fam_genus = cbind(extraction_ncbi_order, extraction_ncbi_family, extraction_ncbi_genus)
species_taxonic = data.frame(species = species_list, extraction_ncbi_order_fam_genus)
head(species_taxonic, 5)
```

* Upload:

```{r solution_upload_dataset, eval = F}
bef.portal.upload.dataset(dataset = species_taxonic, dataset_title = "Taxonomic information for species in ...")
```

---

## Thanks four your attention
